---
title: "affirmoTM tutorial 2: input files"
subtitle: "Each input file: its layout, its rules, and how to look at it"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry: margin=2cm
    fontsize: 10pt
    code-overflow: wrap
    pdf-engine: xelatex
    fig-width: 7
    fig-height: 4.5
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,fontsize=\small}
        \RecustomVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
execute:
  message: false
  warning: true
---

```{r setup}
#| include: false
# Rendering this tutorial: load the package from the repository and work in a fresh, empty analysis folder.
repo <- rprojroot::find_package_root_file()
devtools::load_all(repo, quiet = TRUE)
library(data.table)
analysis <- file.path(tempdir(), "affirmoTM-tutorial-2")
unlink(analysis, recursive = TRUE)
dir.create(analysis)
knitr::opts_knit$set(root.dir = analysis)
knitr::opts_chunk$set(message = FALSE, dev = "cairo_pdf")
if (!isTRUE(getOption("knitr.in.progress"))) setwd(analysis)
options(width = 85)
```

This tutorial starts from a fresh analysis folder with the example input files (see tutorial 1):

```{r start}
#| eval: false
library(affirmoTM)
```

```{r example-inputs}
tm_example_inputs()
```

# The input files

All inputs live in `inputs/`, with fixed names. Your own files must use the same names and layouts.

| File | What it holds | Read with |
|----------------------------------|------------------------------------|--------------------------|
| `settings.yaml` | Run settings (tutorial 3) | `tm_read_settings()` |
| `transition_matrix_usual_care.csv`, `transition_matrix_iabc.csv` | Yearly transition probabilities, one matrix per arm | `tm_read_matrices()` |
| `cohort.csv` | The patients | `tm_read_cohort()` |
| `entry_states.csv` | The chance of each state in a patient's first year | `tm_read_entry_states()` |
| `costs.csv` | The cost of a year in each state, per arm | `tm_read_costs()` |
| `utilities.csv` | The QALY weight of a year in each state, per arm | `tm_read_utilities()` |
| `intervention_cost.csv` | The cost of delivering iABC, by year after entry | `tm_read_intervention_cost()` |

Each reader checks its file and stops with a message naming the file and the problem. `tm_run()` reads and checks
every file before it runs anything.

```{r settings}
settings <- tm_read_settings()
```

# Transition matrices

```{r matrices}
matrices <- tm_read_matrices(settings)
matrices$usual_care
```

The probabilities are in `probs`: one row per stratum (age group, sex, cluster) and current state (`from`), and one
column per state, holding the probability of being in that state next year.

```{r matrices-probs}
usual_care <- matrices$usual_care$probs
states <- matrices$usual_care$states
usual_care[1:3, 1:9]
```

The rules, and a quick way to look at each one:

- **Every row sums to 1.**

  ```{r rule-sum}
  usual_care[, range(rowSums(.SD)), .SDcols = states]
  ```

- **Every stratum has a row for every state** (19 here):

  ```{r rule-complete}
  usual_care[, .N, by = .(agegrp, sex, mm_cluster)][, range(N)]
  ```

- **Every state has its own rows, death states included.** A death state's rows go back to the same state with
  probability 1; that is how the package recognises death states. For example:

  ```{r rule-death}
  usual_care[from == "CurrentDeath", unique(CurrentDeath)]
  ```

- **The stratifier columns are the ones listed under `strata` in `settings.yaml`.** The age-group column must be called
  `agegrp`: it is the only stratifier that changes as patients age. Age groups are written as ranges, without gaps:

  ```{r rule-age}
  unique(usual_care$agegrp)
  ```

- **Both matrices have the same states and the same strata.**
- **The probabilities themselves are your responsibility.** The package checks the layout, not the numbers. When
  `tm_run()` starts, it lists the death states it found; check that no living state is among them.

# Cohort

```{r cohort}
cohort <- tm_read_cohort(settings, matrices)
cohort
```

| Column | Meaning |
|----------------------|------------------------------------------------------|
| `id` | A unique patient id |
| `age` | Age in whole years in the patient's entry cycle (not at cycle 0) |
| `sex`, `mm_cluster` | One column for each stratifier other than `agegrp` |
| `entry_cycle` | The cycle in which the patient enters; 0 is the first year |
| `agegrp` | Optional: if it is missing, it is worked out from `age` |

Patients can enter in different cycles:

```{r cohort-entry}
cohort[, .N, keyby = entry_cycle]
```

The rules:

- ages lie within the matrices' age groups, and every value of `sex` and `mm_cluster` appears in the matrices;
- the matrices have rows for every stratum a patient reaches as they age;
- patients who enter in cycle `horizon` or later are left out, with a warning.

# Entry states

The chance of each state in a patient's entry year, one row per stratum. Both arms start from the same entry states.

```{r entry-states}
entry_states <- tm_read_entry_states(settings, matrices)
entry_states[agegrp == "75-79" & sex == "women" & mm_cluster == "Cardiovascular"]
```

Every row sums to 1, and every patient's entry stratum needs a row:

```{r entry-rule}
entry_states[, range(rowSums(.SD)), .SDcols = setdiff(names(entry_states), settings$strata)]
```

# Costs

The cost of a year spent in each state, one column per arm. Every state needs a row, death states included (the cost
of the year of death). Costs are 0 or more.

```{r costs}
costs <- tm_read_costs(settings, matrices)
costs
```

- **Entry year:** both arms are charged the `usual_care` cost, because both arms are in the same state. The `iabc` cost
  applies from the year after entry.
- **The `iabc` column:** use it for differences in care that depend on the state, such as medication. The cost of
  delivering the programme itself goes in `intervention_cost.csv`.

# Utilities

The QALY weight of a year spent in each living state, one column per arm. Utilities are at most 1. Death states count
0 in the year of death, so they can be left out. In the entry year both arms take the `usual_care` utility.

```{r utilities}
utilities <- tm_read_utilities(settings, matrices)
utilities
```

# Intervention cost

The cost of delivering iABC to one patient in each year after their entry (`year` 1 is the first year after entry),
one row for every year from 1 to `horizon - 1`. It is the same whatever the patient's state, and is charged to iABC
patients who are alive in that year.

```{r intervention}
intervention <- tm_read_intervention_cost(settings)
intervention
```

# Values that change over time: a `year` column

A matrix, `costs.csv` or `utilities.csv` can change with the years since a patient's entry. Add a `year` column (1 is
the first year after entry) and give a full set of rows for every year from 1 to `horizon - 1`. In a matrix, the rows
for year *k* drive the move into year *k* after entry.

For example, an iABC effect that lasts one year: iABC's own probabilities in year 1, and usual care's in years 2 to 9.

```{r year-build}
usual_care_file <- fread("inputs/transition_matrix_usual_care.csv")
iabc_file <- fread("inputs/transition_matrix_iabc.csv")
later_years <- usual_care_file[rep(seq_len(.N), 8)][, year := rep(2:9, each = nrow(usual_care_file))]
iabc_by_year <- rbind(iabc_file[, year := 1L], later_years)
iabc_by_year[, .N, keyby = year]
fwrite(iabc_by_year, "inputs/transition_matrix_iabc.csv")
```

The package reads it as a matrix that changes by year:

```{r year-read}
tm_read_matrices(settings)$iabc
```

The result, next to the example's, with 20 runs each:

```{r year-run}
twenty_runs <- settings
twenty_runs$n_runs <- 20
one_year <- tm_run(twenty_runs)$comparison
tm_example_inputs(overwrite = TRUE)  # put the example files back
every_year <- tm_run(twenty_runs)$comparison
rbind(cbind(effect = "every year", every_year[, .(delta_cost, delta_qaly, icer, quadrant)]),
      cbind(effect = "first year only", one_year[, .(delta_cost, delta_qaly, icer, quadrant)]))
```

# Costs and utilities that differ between patients

`costs.csv` and `utilities.csv` can also have stratifier columns, any of the `strata` in `settings.yaml`. Each
patient-year then uses the patient's stratum in that year: their current age group, as they age, and their sex and
cluster. Every stratum needs a row for every state.

For example, utilities that fall with age and are lower for women, and costs that differ by cluster. The factors are
invented for this example.

```{r stratum-build}
utilities_file <- fread("inputs/utilities.csv")
age_groups <- unique(usual_care$agegrp)
by_age_sex <- utilities_file[, .(agegrp = age_groups), by = .(state, usual_care, iabc)]
by_age_sex <- by_age_sex[, .(sex = c("men", "women")), by = .(state, usual_care, iabc, agegrp)]
by_age_sex[, factor := 1 - 0.03 * (as.integer(sub("-.*", "", agegrp)) - 65) / 5 - 0.02 * (sex == "women")]
by_age_sex[, `:=`(usual_care = usual_care * factor, iabc = iabc * factor)][, factor := NULL]
by_age_sex[state == "NoEvent" & agegrp %in% c("65-69", "85-89")]

costs_file <- fread("inputs/costs.csv")
cluster_factor <- c(Cardiovascular = 1.2, Complex = 1.5, Eye = 1.0, Metabolic = 1.1, MSK = 1.0,
                    Neuropsychiatric = 1.2, Unspecified = 1.0)
by_cluster <- costs_file[, .(mm_cluster = names(cluster_factor)), by = .(state, usual_care, iabc)]
by_cluster[, `:=`(usual_care = usual_care * cluster_factor[mm_cluster], iabc = iabc * cluster_factor[mm_cluster])]
by_cluster[state == "NoEvent"]

fwrite(by_age_sex, "inputs/utilities.csv")
fwrite(by_cluster, "inputs/costs.csv")
```

The result, next to the example's:

```{r stratum-run}
by_stratum <- tm_run(twenty_runs)$comparison
tm_example_inputs(overwrite = TRUE)  # put the example files back
rbind(cbind(values = "the same for everyone", every_year[, .(delta_cost, delta_qaly, icer, quadrant)]),
      cbind(values = "by age, sex and cluster", by_stratum[, .(delta_cost, delta_qaly, icer, quadrant)]))
```
