---
title: "affirmoTM tutorial 3: settings"
subtitle: "Each setting in settings.yaml, and how to change them"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry: margin=2cm
    fontsize: 10pt
    code-overflow: wrap
    pdf-engine: xelatex
    fig-width: 7
    fig-height: 4.5
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,fontsize=\small}
        \RecustomVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
execute:
  message: false
  warning: true
---

```{r setup}
#| include: false
# Rendering this tutorial: load the package from the repository and work in a fresh, empty analysis folder.
repo <- rprojroot::find_package_root_file()
devtools::load_all(repo, quiet = TRUE)
library(data.table)
analysis <- file.path(tempdir(), "affirmoTM-tutorial-3")
unlink(analysis, recursive = TRUE)
dir.create(analysis)
knitr::opts_knit$set(root.dir = analysis)
knitr::opts_chunk$set(message = FALSE, dev = "cairo_pdf")
if (!isTRUE(getOption("knitr.in.progress"))) setwd(analysis)
options(width = 85)
```

This tutorial starts from a fresh analysis folder with the example input files (see tutorial 1):

```{r start}
#| eval: false
library(affirmoTM)
```

```{r example-inputs}
tm_example_inputs()
```

# The settings file

`inputs/settings.yaml` holds the run settings, one per line, with comments:

```{r settings-file}
cat(readLines("inputs/settings.yaml"), sep = "\n")
```

| Setting | Example | What it does |
|--------------------------|--------------------|------------------------------------------------------|
| `strata` | `[agegrp, sex, mm_cluster]` | The stratifier columns of the input files. Every other matrix column, apart from `from`, is a state. `[]` for none |
| `seed` | `2026` | The starting seed. Run *r* uses `seed + r - 1`, so results can be repeated exactly |
| `horizon` | `10` | The number of yearly cycles, counted from cycle 0 for everyone |
| `age_limit_death_state` | `CurrentDeath` | The death state of patients who would pass the oldest age group |
| `discount_rate` | `0.03` | The yearly discount rate for costs and QALYs, as a proportion (`0.03` for 3%). A year in cycle *t* counts 1 / (1 + rate)^*t* |
| `n_runs` | `100` | The number of runs; each run is one point on the CE plane |
| `events` | lists of states | The states counted as each event in the results |
| `currency` | `"€"` | The currency symbol in the printed results and the plot |
| `wtp` | `[20000, 30000, 50000]` | Willingness-to-pay thresholds per QALY, drawn as lines on the CE plane. `[]` for none |

`tm_read_settings()` reads the file and checks every setting. A wrong type or value stops with a message naming the
setting, and a name the package does not know, such as a typo, gives a warning.

In R, the settings are a list:

```{r read-settings}
settings <- tm_read_settings()
str(settings)
```

# Changing settings

There are two ways:

- **Edit `inputs/settings.yaml`** and run `tm_run()` again. This is the way to change the analysis itself.
- **Change the list in R** and pass it to `tm_run()`, which is handy for trying things out; the file stays as it is.

The examples below change the list in R, with 20 runs each to keep them quick. Every call to `tm_run()` overwrites
`outputs/`.

```{r base}
base <- tm_read_settings()
base$n_runs <- 20
base_results <- tm_run(base)
```

## Discount rate

```{r discount}
no_discounting <- base
no_discounting$discount_rate <- 0
no_discounting_results <- tm_run(no_discounting)
```

## Time horizon

The example cohort enters between cycles 0 and 9. With a 5-year horizon, the patients who would enter in cycles 5 to 9
fall outside it, and `tm_run()` leaves them out with a warning:

```{r horizon}
five_years <- base
five_years$horizon <- 5
five_years_results <- tm_run(five_years)
```

`intervention_cost.csv` lists years 1 to 9, which covers a 5-year horizon too: later years are allowed and not used.

## Side by side

```{r compare}
rbind(cbind(setting = "3% discount, 10 years", base_results$comparison),
      cbind(setting = "no discounting, 10 years", no_discounting_results$comparison),
      cbind(setting = "3% discount, 5 years", five_years_results$comparison)
)[, .(setting, delta_cost, delta_qaly, icer, quadrant)]
```

## Willingness-to-pay lines

`wtp` only changes the lines on the CE plane:

```{r wtp}
other_lines <- base
other_lines$wtp <- c(10000, 25000)
plot(tm_run(other_lines))
```

## Seed and number of runs

The same seed always gives the same results; another seed gives another set of runs:

```{r seed}
two_runs <- base
two_runs$n_runs <- 2
first <- tm_run(two_runs)$comparison
again <- tm_run(two_runs)$comparison
two_runs$seed <- 1
other_seed <- tm_run(two_runs)$comparison
rbind(cbind(run = "seed 2026", first), cbind(run = "seed 2026 again", again),
      cbind(run = "seed 1", other_seed))[, .(run, delta_cost, delta_qaly)]
```

More runs give more points on the CE plane and a steadier mean. Each run of the 5,000 example patients takes about
a quarter of a second.
