---
title: "affirmoTM tutorial 4: running the model"
subtitle: "What tm_run() does, and the steps of a run one at a time"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry: margin=2cm
    fontsize: 10pt
    code-overflow: wrap
    pdf-engine: xelatex
    fig-width: 7
    fig-height: 4.5
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,fontsize=\small}
        \RecustomVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
execute:
  message: false
  warning: true
---

```{r setup}
#| include: false
# Rendering this tutorial: load the package from the repository and work in a fresh, empty analysis folder.
repo <- rprojroot::find_package_root_file()
devtools::load_all(repo, quiet = TRUE)
library(data.table)
analysis <- file.path(tempdir(), "affirmoTM-tutorial-4")
unlink(analysis, recursive = TRUE)
dir.create(analysis)
knitr::opts_knit$set(root.dir = analysis)
knitr::opts_chunk$set(message = FALSE, dev = "cairo_pdf")
if (!isTRUE(getOption("knitr.in.progress"))) setwd(analysis)
options(width = 85)
```

This tutorial starts from a fresh analysis folder with the example input files (see tutorial 1):

```{r start}
#| eval: false
library(affirmoTM)
```

```{r example-inputs}
tm_example_inputs()
```

# What `tm_run()` does

`tm_run()` runs the whole model with the files in `inputs/`:

1. reads and checks every input file and setting; any problem stops it before anything is run or written;
2. lists the death states it found in the matrices;
3. runs the model `n_runs` times, run *r* with seed `seed + r - 1`;
4. averages the runs, compares the arms, and writes the results to `outputs/`.

While it runs, it reports the death states and its progress (here with 20 runs):

```{r run-messages}
#| message: true
settings <- tm_read_settings()
settings$n_runs <- 20
results <- tm_run(settings)
```

Check that no living state is among the death states.

## Run time

The run time grows with the number of runs and the number of patients:

```{r run-time}
system.time(tm_run(settings))
```

## The same seed, the same results

```{r reproducible}
again <- tm_run(settings)
identical(results$by_run, again$by_run)
```

## Outputs

Every call writes, and overwrites, the files in `outputs/`:

```{r outputs}
list.files("outputs")
```

Tutorial 5 explains the results.

# The steps of a run, one at a time

`tm_run()` calls the functions below, once per run. They can also be used one at a time, for example to look at one
run in detail. Each takes the `settings` list. Tutorial 6 follows a run step by step in more detail.

**Read the inputs.** Each reader checks its file:

```{r read}
matrices <- tm_read_matrices(settings)
cohort <- tm_read_cohort(settings, matrices)
entry_states <- tm_read_entry_states(settings, matrices)
costs <- tm_read_costs(settings, matrices)
utilities <- tm_read_utilities(settings, matrices)
intervention <- tm_read_intervention_cost(settings)
death <- matrices$usual_care$death
```

**Entry states.** `tm_assign_entry_states()` draws each patient's state in their entry year. Both arms start from
these states:

```{r assign}
patients <- tm_assign_entry_states(cohort, entry_states, settings, seed = settings$seed)
patients[, .N, keyby = entry_state]
```

**Simulation.** `tm_simulate()` moves each patient through both arms, one year at a time, until they die or the
horizon ends. One row per patient, arm and year:

```{r simulate}
sim <- tm_simulate(patients, matrices, settings, seed = settings$seed)
sim[id == 1]
```

**Costs and QALYs.** `tm_add_outcomes()` gives each year its cost, intervention cost, QALY and discount factor:

```{r outcomes}
out <- tm_add_outcomes(sim, costs, utilities, intervention, death, settings)
out[id == 1, .(arm, cycle, state, cost, intervention_cost, qaly, discount)]
```

**Summary.** `tm_summarise()` adds up each arm:

```{r summarise}
one_run <- tm_summarise(out, death, settings)
one_run[, .(arm, patients, cost_per_patient, qaly_per_patient, deaths)]
```

**Comparison.** `tm_compare()` takes iABC minus usual care:

```{r compare}
tm_compare(one_run[, `:=`(run = 1L, seed = settings$seed)])$overall
```

This is exactly the first run of `tm_run()`, which uses the same seed:

```{r same-as-run-1}
results$by_run[run == 1, .(arm, patients, cost_per_patient, qaly_per_patient, deaths)]
```
