---
title: "affirmoTM tutorial 5: results"
subtitle: "The results object, the comparison, the CE plane and the output files"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry: margin=2cm
    fontsize: 10pt
    code-overflow: wrap
    pdf-engine: xelatex
    fig-width: 7
    fig-height: 4.5
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,fontsize=\small}
        \RecustomVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
execute:
  message: false
  warning: true
---

```{r setup}
#| include: false
# Rendering this tutorial: load the package from the repository and work in a fresh, empty analysis folder.
repo <- rprojroot::find_package_root_file()
devtools::load_all(repo, quiet = TRUE)
library(data.table)
analysis <- file.path(tempdir(), "affirmoTM-tutorial-5")
unlink(analysis, recursive = TRUE)
dir.create(analysis)
knitr::opts_knit$set(root.dir = analysis)
knitr::opts_chunk$set(message = FALSE, dev = "cairo_pdf")
if (!isTRUE(getOption("knitr.in.progress"))) setwd(analysis)
options(width = 85)
```

This tutorial starts from a fresh analysis folder with the example input files, and runs the model (see tutorial 1):

```{r start}
#| eval: false
library(affirmoTM)
```

```{r run}
tm_example_inputs()
results <- tm_run()
```

# The results object

`tm_run()` returns the results and prints a summary:

```{r print}
results
```

The results are a list:

```{r names}
names(results)
```

| Element | What it holds |
|------------------------------|--------------------------------------------------|
| `by_run` | One row per run and arm: the summary of each run |
| `summary` | The mean over the runs, one row per arm |
| `comparison_by_run` | iABC minus usual care, one row per run |
| `comparison` | The mean differences, the ICER and its quadrant |
| `n_runs`, `currency`, `wtp` | The settings used |

# Each arm

```{r summary}
results$summary
```

| Column | Meaning |
|------------------------------------|--------------------------------------------------|
| `patients` | Patients who enter within the horizon |
| `cost_total`, `qaly_total` | Discounted totals for the whole cohort |
| `intervention_cost_total` | The discounted intervention cost within `cost_total` |
| `cost_per_patient` | `cost_total` divided by the number of patients |
| `qaly_per_patient` | `qaly_total` divided by the number of patients |
| `intervention_cost_per_patient` | `intervention_cost_total` divided by the number of patients |
| one column per event | The years spent in that event's states (see `events` in `settings.yaml`) |
| `deaths` | The number of patients who died |

Costs and QALYs are per patient; events and deaths are counts for the whole cohort.

# The comparison

iABC minus usual care, as the mean over the runs:

```{r comparison}
results$comparison
```

- `delta_cost` and `delta_qaly`: the incremental cost and QALYs per patient.
- `delta_intervention_cost`: the part of `delta_cost` that is intervention cost.
- the other `delta_` columns: the difference in each event count and in deaths, for the whole cohort.
- `icer`: the mean incremental cost divided by the mean incremental QALYs.
- `quadrant`: where the result lies on the CE plane. It says how to read the ICER:

| `quadrant` | iABC compared with usual care | Printed as |
|--------------------------|----------------------------------|----------------------|
| `costlier, more QALYs` | Costs more, gives more QALYs | € ... per QALY gained |
| `dominant` | Costs less (or the same), gives more QALYs | iABC dominant |
| `dominated` | Costs more (or the same), gives fewer QALYs | iABC dominated |
| `cheaper, fewer QALYs` | Costs less, gives fewer QALYs | € ... saved per QALY lost |
| `no QALY difference` | The same QALYs | undefined |


The same differences for each run:

```{r by-run}
head(results$comparison_by_run)
results$comparison_by_run[, .(mean = mean(delta_qaly), lowest = min(delta_qaly), highest = max(delta_qaly))]
```

# The cost-effectiveness plane

```{r plane}
plot(results)
```

- Each blue point is one run: incremental QALYs per patient across, incremental cost per patient up.
- The diamond is the mean across the runs, labelled with the result in the words of its quadrant.
- The dashed lines are the willingness-to-pay thresholds in `wtp`. Points below a line are cost-effective at that
  threshold.
- The runs use the same inputs and differ only in their random numbers, so the spread of the points shows how much
  a single run varies by chance.

`plot(results)` returns the plot, so it can be saved or changed with ggplot2:

```{r save-plot}
#| eval: false
p <- plot(results)
ggplot2::ggsave("my_ce_plane.pdf", p, width = 7, height = 5)
```

# The output files

Every call to `tm_run()` writes these files to `outputs/`:

| File | What it holds |
|----------------------------|--------------------------------------------------|
| `summary_by_run.csv` | One row per run and arm (`results$by_run`) |
| `summary.csv` | The mean per arm (`results$summary`) |
| `comparison_by_run.csv` | iABC minus usual care, one row per run (`results$comparison_by_run`) |
| `comparison.csv` | The mean differences, the ICER and its quadrant (`results$comparison`) |
| `ce_plane.png` | The CE plane |

```{r files}
list.files("outputs")
fread("outputs/comparison.csv")
```
