---
title: "affirmoTM tutorial 6: walkthrough"
subtitle: "One run followed in detail: patients, costs, QALYs and how the arms behave over time"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    geometry: margin=2cm
    fontsize: 10pt
    code-overflow: wrap
    pdf-engine: xelatex
    fig-width: 7
    fig-height: 4.5
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere,fontsize=\small}
        \RecustomVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
execute:
  message: false
  warning: true
---

```{r setup}
#| include: false
# Rendering this tutorial: load the package from the repository and work in a fresh, empty analysis folder.
repo <- rprojroot::find_package_root_file()
devtools::load_all(repo, quiet = TRUE)
library(data.table)
library(ggplot2)
analysis <- file.path(tempdir(), "affirmoTM-tutorial-6")
unlink(analysis, recursive = TRUE)
dir.create(analysis)
knitr::opts_knit$set(root.dir = analysis)
knitr::opts_chunk$set(message = FALSE, dev = "cairo_pdf")
if (!isTRUE(getOption("knitr.in.progress"))) setwd(analysis)
options(width = 85)
```

# About this tutorial

This tutorial follows the example inputs through the model: what goes in, what each function produces, and how the
two arms behave over time.

- **Inputs:** each one is read with the package's own reader and explored with short `data.table` commands.
- **One run, followed step by step:** a few patients are followed year by year, with their costs and QALYs worked out
  by hand.
- **Over time:** the cohort is followed across the horizon in both arms.
- **The full model:** `tm_run()` at the end.

It starts from a fresh analysis folder with the example input files (see tutorial 1):

```{r start}
#| eval: false
library(affirmoTM)
```

```{r example-inputs}
tm_example_inputs()
```

# The inputs

## Settings

```{r settings}
settings <- tm_read_settings()
str(settings)
```

## Transition matrices

`tm_read_matrices()` reads both matrices and checks them. Each one is a `tm_matrix`; printing it gives an overview:

```{r matrices}
matrices <- tm_read_matrices(settings)
matrices$usual_care
```

The probabilities are in `probs`, one row per stratum and current state (`from`), and one column per next state:

```{r matrices-probs}
usual_care <- matrices$usual_care$probs
iabc <- matrices$iabc$probs
states <- matrices$usual_care$states
death <- matrices$usual_care$death
strata <- settings$strata
age_order <- unique(usual_care$agegrp)[order(as.integer(sub("-.*", "", unique(usual_care$agegrp))))]  # for plots
arm_colours <- c(usual_care = "grey35", iabc = "#2a78d6")
dim(usual_care)
usual_care[1:3, 1:10]
```

Every row sums to 1, in both matrices:

```{r rows-sum}
usual_care[, .(min = min(rowSums(.SD)), max = max(rowSums(.SD))), .SDcols = states]
iabc[, .(min = min(rowSums(.SD)), max = max(rowSums(.SD))), .SDcols = states]
```

The matrices are easier to explore in long form, one row per move, with both arms side by side:

```{r long}
long_usual_care <- melt(usual_care, id.vars = c(strata, "from"), variable.name = "to", value.name = "p",
                        variable.factor = FALSE)
long_iabc <- melt(iabc, id.vars = c(strata, "from"), variable.name = "to", value.name = "p", variable.factor = FALSE)
moves <- long_usual_care[long_iabc, on = c(strata, "from", "to"),
                         .(agegrp, sex, mm_cluster, from, to, usual_care = p, iabc = i.p)]
moves
```

A death state's rows always go back to the same state. In every stratum, the only move out of a death state is to
itself, with probability 1:

```{r death-rows}
moves[from %in% death & usual_care > 0, .(rows = .N, probability = unique(usual_care)), by = .(from, to)]
```

The moves that are possible from each state, i.e. with a probability above 0 in at least one stratum:

```{r possible-moves}
knitr::kable(moves[usual_care > 0, .(possible_next_states = paste(unique(to), collapse = ", ")), by = from])
```

The same patient under each arm: a man aged 75–79 in the Unspecified cluster, without an event this year. Each number
is the probability of being in that state next year:

```{r one-patient-row}
moves[agegrp == "75-79" & sex == "men" & mm_cluster == "Unspecified" & from == "NoEvent" & usual_care > 0,
      .(to, usual_care, iabc, difference = iabc - usual_care)]
```

The yearly probability of death for patients without an event, by age group, sex and cluster:

```{r death-by-age}
#| fig-height: 5.5
death_by_age <- moves[from == "NoEvent" & to %in% death, .(death = sum(usual_care)), by = .(agegrp, sex, mm_cluster)]
ggplot(death_by_age, aes(factor(agegrp, levels = age_order), death, colour = sex, group = sex)) +
  geom_line() + geom_point(size = 1) +
  facet_wrap(~mm_cluster, ncol = 4) +
  labs(x = "Age group", y = "Probability of death within a year", colour = NULL) +
  theme_minimal(base_size = 10) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5), legend.position = "top")
```

The yearly probability of a stroke for patients without an event, under each arm, for the Unspecified cluster:

```{r stroke-by-age}
stroke_states <- unique(unlist(settings$events[c("ischaemic_stroke", "haemorrhagic_stroke")]))
stroke_by_age <- moves[from == "NoEvent" & to %in% stroke_states & mm_cluster == "Unspecified",
                       .(usual_care = sum(usual_care), iabc = sum(iabc)), by = .(agegrp, sex)]
ggplot(melt(stroke_by_age, id.vars = c("agegrp", "sex"), variable.name = "arm", value.name = "stroke"),
       aes(factor(agegrp, levels = age_order), stroke, colour = arm, group = arm)) +
  geom_line() + geom_point(size = 1) + facet_wrap(~sex) + scale_colour_manual(values = arm_colours, breaks = names(arm_colours)) +
  labs(x = "Age group", y = "Probability of a stroke within a year", colour = NULL) +
  theme_minimal(base_size = 10) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5), legend.position = "top")
```

## Cohort

`tm_read_cohort()` reads the patients and works out each patient's age group from their age:

```{r cohort}
cohort <- tm_read_cohort(settings, matrices)
cohort
```

Patients enter in different cycles. Cycle 0 (2013) holds everyone already living with AF when the data start, and
later cycles hold new patients:

```{r cohort-entry}
cohort[, .(patients = .N, mean_age = round(mean(age), 1)), keyby = entry_cycle]
```

```{r cohort-groups}
cohort[, .N, keyby = sex]
cohort[, .N, keyby = mm_cluster]
```

```{r cohort-age}
ggplot(cohort, aes(age)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white") +
  facet_wrap(~sex) +
  labs(x = "Age at entry", y = "Patients") +
  theme_minimal(base_size = 10)
```

## Entry states

`tm_read_entry_states()` reads the probability of each state in a patient's entry year, one row per stratum. Every row
sums to 1:

```{r entry-states}
entry_states <- tm_read_entry_states(settings, matrices)
entry_cols <- setdiff(names(entry_states), strata)
entry_states[, .(rows = .N, min = min(rowSums(.SD)), max = max(rowSums(.SD))), .SDcols = entry_cols]
entry_states[agegrp == "75-79" & sex == "women" & mm_cluster == "Cardiovascular"]
```

## Costs, utilities and the intervention cost

```{r values}
costs <- tm_read_costs(settings, matrices)
utilities <- tm_read_utilities(settings, matrices)
intervention <- tm_read_intervention_cost(settings)
costs
utilities
intervention
```

- **Costs:** the cost of a year spent in each state, per arm.
- **Utilities:** the QALY weight of a year in each state, per arm. Death states count 0.
- **Intervention cost:** the cost of delivering iABC to one patient in each year after entry. It is charged to iABC
  patients who are alive in that year.

# One run, step by step

`tm_run()` repeats these steps once per run. Here they are done one at a time for the first run, which uses the seed
from `settings.yaml`:

```{r seed}
seed <- settings$seed
seed
```

## Entry states

`tm_assign_entry_states()` draws each patient's state in their entry year from their row of the entry-state table. Both
arms start from these states.

```{r assign}
patients <- tm_assign_entry_states(cohort, entry_states, settings, seed = seed)
patients[, .N, keyby = entry_state][order(-N)]
```

## The simulation

`tm_simulate()` moves every patient through both arms one year at a time. In each later year, it draws the patient's
next state from their matrix row: their arm, sex and cluster, their age group in the previous year, and their previous
state. The result has one row per patient, arm and year, from the entry year up to and including the year of death:

```{r simulate}
sim <- tm_simulate(patients, matrices, settings, seed = seed)
setorder(sim, arm, id, cycle)
sim
sim[, .(patients = uniqueN(id), patient_years = .N, deaths = sum(state %in% death)), by = arm]
```

## Following five patients

Each patient's path, written as one line per arm, makes it easy to pick patients with different stories:

```{r paths}
paths <- sim[, .(path = paste(state, collapse = " > "), entered = min(cycle), last_state = last(state)), by = .(arm, id)]
paths <- dcast(paths, id + entered ~ arm, value.var = c("path", "last_state"))
head(paths[, .(id, entered, path_usual_care)])
```

The five patients, chosen from these paths:

```{r chosen}
event_free <- paths[entered == 0 & path_usual_care == paste(rep("NoEvent", 10), collapse = " > "), id][1]
stroke <- paths[entered == 0 & grepl("NoEvent > AIS > PIS", path_usual_care), id][1]
died <- paths[entered == 0 & last_state_usual_care == "CurrentDeath" & grepl("^NoEvent > NoEvent > NoEvent", path_usual_care), id][1]
late <- paths[entered == 7 & path_usual_care == path_iabc, id][1]
differs <- paths[grepl("AIS", path_usual_care) & !grepl("AIS", path_iabc), id][1]
c(event_free = event_free, stroke = stroke, died = died, late = late, differs = differs)
```

Each patient's years in both arms, side by side:

**A patient who stays event-free** for the whole horizon:

```{r patient-event-free}
dcast(sim[id == event_free], cycle + age + agegrp ~ arm, value.var = "state")
```

**A patient with an ischaemic stroke:** acute (`AIS`) in the year it happens, then past (`PIS`) in the years after:

```{r patient-stroke}
dcast(sim[id == stroke], cycle + age + agegrp ~ arm, value.var = "state")
```

**A patient who dies.** The year of death is the last year recorded:

```{r patient-died}
dcast(sim[id == died], cycle + age + agegrp ~ arm, value.var = "state")
```

**A patient who enters late**, in cycle 7 (2020), and is followed until the end of the horizon:

```{r patient-late}
dcast(sim[id == late], cycle + age + agegrp ~ arm, value.var = "state")
```

**A patient whose path differs between the arms.** Both arms use the same random number for the same patient and year,
so the paths stay together until, in one year, the same number lands on a different state in each arm's row:

```{r patient-differs}
dcast(sim[id == differs], cycle + age + agegrp ~ arm, value.var = "state")
```

## Their costs and QALYs, by hand

`tm_add_outcomes()` gives each patient-year its cost, intervention cost, QALY and discount factor:

```{r outcomes}
out <- tm_add_outcomes(sim, costs, utilities, intervention, death, settings)
out[id == differs, .(arm, cycle, state, cost, intervention_cost, qaly, discount)]
```

The same numbers worked out by hand for the five patients, from the input tables:

- **Values:** in a patient's entry year both arms take the usual-care cost and utility; from the next year, each arm's
  own.
- **Intervention cost:** iABC patients pay the intervention cost of that year since entry, for each year after entry
  that they are alive.
- **Discount factor:** a year in cycle *t* is discounted by 1 / (1 + rate)^*t*.

```{r by-hand}
costs_long <- melt(costs, id.vars = "state", variable.name = "arm", value.name = "state_cost", variable.factor = FALSE)
utilities_long <- melt(utilities, id.vars = "state", variable.name = "arm", value.name = "utility",
                       variable.factor = FALSE)
followed <- c(event_free, stroke, died, late, differs)
by_hand <- out[id %in% followed, .(id, arm, cycle, state, cost, intervention_cost, qaly, discount)]
by_hand[, years_after_entry := cycle - min(cycle), by = id]
by_hand[, value_arm := fifelse(years_after_entry == 0, "usual_care", arm)]
by_hand[costs_long, on = .(state, value_arm = arm), state_cost := i.state_cost]
by_hand[utilities_long, on = .(state, value_arm = arm), utility := i.utility]
by_hand[, programme := fifelse(arm == "iabc" & years_after_entry > 0 & !state %in% death,
                               intervention$cost[match(years_after_entry, intervention$year)], 0)]
by_hand[, `:=`(cost_by_hand = state_cost + programme, discount_by_hand = 1 / (1 + settings$discount_rate)^cycle)]
by_hand[id == differs, .(arm, cycle, state, state_cost, programme, cost_by_hand, cost)]
by_hand[id == differs, .(arm, cycle, state, utility, qaly, discount_by_hand = round(discount_by_hand, 4),
                         discount = round(discount, 4))]
```

Every year of the five patients, compared with the package's numbers:

```{r by-hand-compare}
by_hand[, .(years = .N, same_cost = all(cost == cost_by_hand), same_intervention_cost = all(intervention_cost == programme),
            same_qaly = all(qaly == utility), same_discount = all(abs(discount - discount_by_hand) < 1e-12)), by = id]
```

Each patient's discounted total in each arm, which is what the arm totals add up:

```{r by-hand-totals}
dcast(by_hand[, .(cost = sum(cost_by_hand * discount_by_hand), qalys = sum(utility * discount_by_hand)), by = .(id, arm)],
      id ~ arm, value.var = c("cost", "qalys"))
```

# How the model behaves over time

## States over time

To see the cohort move through the states, the next plots follow the 2,664 patients who enter in cycle 0, so later
entrants don't blur the picture. The 19 states are grouped into five:

```{r state-groups}
state_groups <- data.table(state = states)
state_groups[, group := fcase(state == "NoEvent", "No event",
                              state %in% death, "Dead",
                              state %in% c("AIS", "AHS", "AIS_AB", "AHS_AB", "PB_AIS", "PB_AHS"), "Stroke this year",
                              state %in% c("AB", "PS_AB"), "Bleed this year",
                              default = "Past stroke or bleed")]
state_groups[, .(states = paste(state, collapse = ", ")), by = group]
```

Each year, the share of these patients in each group. A patient who has died stays counted as dead in later years.
The patients already dead in cycle 0 are those whose entry state is a death state: they died in their first AF year.
Both arms start from the same entry states, so they are the same in both arms:

```{r states-over-time}
first_entrants <- cohort[entry_cycle == 0, id]
yearly <- sim[id %in% first_entrants][state_groups, on = "state", nomatch = NULL][, .N, by = .(arm, cycle, group)]
# Years with nobody in a group count 0.
yearly <- yearly[CJ(arm = c("usual_care", "iabc"), cycle = 0:(settings$horizon - 1), group = unique(state_groups$group)),
                 on = .(arm, cycle, group)][is.na(N), N := 0L]
alive <- yearly[group != "Dead"]
dead <- yearly[group == "Dead"][order(cycle), .(cycle, N = cumsum(N), group = "Dead"), by = arm]
over_time <- rbind(alive, dead)[, share := N / length(first_entrants)]
over_time[, `:=`(group = factor(group, levels = c("Dead", "Stroke this year", "Bleed this year", "Past stroke or bleed",
                                                  "No event")), arm = factor(arm, levels = c("usual_care", "iabc")))]
ggplot(over_time, aes(cycle, share, fill = group)) +
  geom_area(stat = "identity") + facet_wrap(~arm) +
  scale_x_continuous(breaks = 0:9) + scale_y_continuous(labels = scales::percent) +
  labs(x = "Cycle (year)", y = "Patients who entered in cycle 0", fill = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "top")
dcast(over_time[cycle %in% c(0, 5, 9)], arm + group ~ paste("cycle", cycle), value.var = "share")
```

## Survival

The share of these patients still alive at the end of each year, and the deaths each year in each arm. The two
lines lie almost on top of each other: in this run, iABC changes the number of deaths among these 2,664 patients by
only a few, as the table shows:

```{r survival}
survival <- dead[, .(arm, cycle, alive = 1 - N / length(first_entrants))]
ggplot(survival, aes(cycle, alive, colour = arm)) +
  geom_line() + geom_point(size = 1) + scale_colour_manual(values = arm_colours, breaks = names(arm_colours)) +
  scale_x_continuous(breaks = 0:9) + scale_y_continuous(labels = scales::percent) +
  labs(x = "Cycle (year)", y = "Still alive", colour = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "top")
deaths_by_year <- dcast(yearly[group == "Dead"], cycle ~ arm, value.var = "N")
deaths_by_year[, `:=`(difference = iabc - usual_care, cumulative_difference = cumsum(iabc - usual_care))]
deaths_by_year
```

## Strokes and bleeds per year

The events counted in the results (`events` in `settings.yaml`), each year, for the whole cohort. Each year a patient
spends in one of an event's states counts as one event:

```{r events}
event_states <- setnames(as.data.table(stack(lapply(settings$events, unlist))), c("state", "event"))
events_by_year <- sim[event_states, on = "state", allow.cartesian = TRUE, nomatch = NULL][, .N, by = .(arm, cycle, event)]
ggplot(events_by_year, aes(cycle, N, colour = arm)) +
  geom_line() + geom_point(size = 1) + facet_wrap(~event, scales = "free_y") + scale_colour_manual(values = arm_colours, breaks = names(arm_colours)) +
  scale_x_continuous(breaks = 0:9) +
  labs(x = "Cycle (year)", y = "Events in the cohort", colour = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "top")
dcast(events_by_year, event ~ arm, value.var = "N", fun.aggregate = sum)[, difference := iabc - usual_care][]
```

## Costs and QALYs building up

The discounted cost and QALYs per patient, added up year by year, for the whole cohort. The cost is split into the
intervention cost and the cost of care (the state costs):

```{r cumulative}
n_patients <- uniqueN(out$id)
by_year <- out[, .(care_cost = sum((cost - intervention_cost) * discount), intervention_cost = sum(intervention_cost * discount),
                   qalys = sum(qaly_disc)), by = .(arm, cycle)]
setorder(by_year, arm, cycle)
by_year[, `:=`(care_cost = cumsum(care_cost) / n_patients, intervention_cost = cumsum(intervention_cost) / n_patients,
               qalys = cumsum(qalys) / n_patients), by = arm]
difference <- dcast(by_year, cycle ~ arm, value.var = c("care_cost", "intervention_cost", "qalys"))
difference <- difference[, .(cycle, care_cost = care_cost_iabc - care_cost_usual_care,
                             intervention_cost = intervention_cost_iabc - intervention_cost_usual_care,
                             total_cost = care_cost_iabc + intervention_cost_iabc - care_cost_usual_care - intervention_cost_usual_care,
                             qalys = qalys_iabc - qalys_usual_care)]
difference
```

iABC minus usual care, added up year by year. The intervention cost is paid in the first year after each patient's
entry, so it rises most in cycle 1 and then a little each year as later entrants reach their first year. The care
costs saved through avoided strokes and bleeds build up over the years:

```{r cumulative-plot}
cost_parts <- melt(difference, id.vars = "cycle", measure.vars = c("care_cost", "intervention_cost", "total_cost"))
cost_parts[, variable := factor(variable, labels = c("Care costs", "Intervention cost", "Total"))]
ggplot(cost_parts, aes(cycle, value, colour = variable)) +
  geom_hline(yintercept = 0, colour = "grey70") + geom_line() + geom_point(size = 1) +
  scale_x_continuous(breaks = 0:9) +
  labs(x = "Cycle (year)", y = "iABC minus usual care, per patient (€, discounted)", colour = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "top")
ggplot(difference, aes(cycle, qalys)) +
  geom_hline(yintercept = 0, colour = "grey70") + geom_line() + geom_point(size = 1) +
  scale_x_continuous(breaks = 0:9) +
  labs(x = "Cycle (year)", y = "iABC minus usual care, QALYs per patient (discounted)") +
  theme_minimal(base_size = 10)
```

## The run's summary and comparison

`tm_summarise()` adds up each arm, and `tm_compare()` takes iABC minus usual care:

```{r summarise}
one_run <- tm_summarise(out, death, settings)
one_run
compared <- tm_compare(copy(one_run)[, `:=`(run = 1L, seed = seed)])
compared$overall
```

The incremental cost and QALYs are the last row of the year-by-year differences above:

```{r summarise-link}
difference[.N, .(total_cost, qalys)]
```

## Subgroups

The discounted cost and QALYs of each patient, and the difference between the arms, by sex, by age group at entry
and by cluster:

```{r subgroups}
per_patient <- out[, .(cost = sum(cost_disc), qalys = sum(qaly_disc)), by = .(id, arm)][cohort, on = "id"]
per_patient <- dcast(per_patient, id + sex + agegrp + mm_cluster ~ arm, value.var = c("cost", "qalys"))
per_patient[, `:=`(delta_cost = cost_iabc - cost_usual_care, delta_qalys = qalys_iabc - qalys_usual_care)]
per_patient[, .(patients = .N, qalys_usual_care = mean(qalys_usual_care), delta_cost = mean(delta_cost),
                delta_qalys = mean(delta_qalys)), keyby = sex]
per_patient[, .(patients = .N, qalys_usual_care = mean(qalys_usual_care), delta_cost = mean(delta_cost),
                delta_qalys = mean(delta_qalys)), keyby = agegrp]
by_cluster <- per_patient[, .(patients = .N, qalys_usual_care = mean(qalys_usual_care), delta_cost = mean(delta_cost),
                              delta_qalys = mean(delta_qalys)), keyby = mm_cluster]
by_cluster
```

```{r subgroups-plot}
ggplot(by_cluster, aes(reorder(mm_cluster, delta_qalys), delta_qalys)) +
  geom_col(fill = "grey60") + geom_text(aes(label = paste(patients, "patients")), hjust = -0.1, size = 3) +
  coord_flip() + expand_limits(y = max(by_cluster$delta_qalys) * 1.4) +
  labs(x = NULL, y = "iABC minus usual care, QALYs per patient (discounted)") +
  theme_minimal(base_size = 10)
```

These come from one run. In small groups, a handful of avoided events moves the average a lot, so subgroup results
from a single run vary more than the overall result.

# The full model: `tm_run()`

`tm_run()` does everything above `n_runs` times. Run *r* uses seed `seed + r − 1`, so its first run is the run followed
in this tutorial. It returns the mean over runs and writes the results to `outputs/`:

```{r run}
results <- tm_run()
results
```

The first run of `tm_run()` and the run followed above, side by side:

```{r run-first}
rbind(cbind(source = "step by step", one_run[, .(arm, cost_per_patient, qaly_per_patient, deaths)]),
      cbind(source = "tm_run(), run 1", results$by_run[run == 1, .(arm, cost_per_patient, qaly_per_patient, deaths)]))
```

Each point on the cost-effectiveness plane is one run; the diamond is the mean across runs:

```{r run-plot}
plot(results)
```

```{r run-files}
list.files("outputs")
```
